---
title: "Figure 2"
author: "Brittany A. Petros"
date: "2024-02-26"
output: pdf_document
---

Set up the workspace.

```{r}
rm(list=ls(all.names=TRUE))
setwd("/Users/bpetros/Desktop/PHIS")
libs <- c("cowplot", "ggsignif", "tidyverse")
invisible(lapply(libs, function(x) suppressPackageStartupMessages(library(x, character.only = TRUE))))
options(stringsAsFactors=FALSE, scipen = 999)
theme_set(theme_classic())
```

Read in the cleaned input files.

```{r}
# read in cleaned pt data
pt <- read.csv("cleaned/rsv_patient.csv")
pt$Date = ymd(pt$Discharge_Date)

# read in cleaned tested pt data
ptest <- read.csv("cleaned/tested_patient.csv")
ptest$Date = ymd(ptest$Discharge_Date)
```

This function takes a df with column "Admit_Age_In_Days" (integer) as input. It creates a column called Age_Group that labels rows as belonging to one of the following age groups: 0-3 months, 3-6 months, 6-12 months, 1-2 years, 2-4 years, and >= 5 years.

```{r}
add_age_group <- function(df) {
  df$Age_Group <- cut(
    df$Admit_Age_In_Days,
    breaks = c(0, 90, 180, 365, 730, 1460, Inf),
    labels = c("0-3 months", "3-6 months", "6-12 months", "1-2 years", "2-4 years", "5-17 years"),
    right = FALSE)
  return(df)}
```

This function takes a df with column "column_name" as input. The column must represent a binary variable using either 0s and 1s (integer) or "N" and "Y" (character). The function prints the percent of rows for which the column equals either 1 or "Y".

```{r}
summarize_and_print_column <- function(df, column_name) {
  df_summary <- df %>%
    summarize(
      Percent = round(sum(.data[[column_name]] == "Y" | .data[[column_name]] == 1) / n() * 100, 2))
  cat(column_name, "- Percent:", df_summary$Percent, "%\n")}
```

This function takes a df generated from a PHIS database download as input (i.e., multiple specific column names are required). It calculates and prints demographic (e.g., age, race, ethnicity) and clinical (e.g., presence of comorbid conditions, outcomes) summary statistics.

```{r}
analyze_dataframe <- function(df) {
  
  # counts and percent of patients in each category (ED, ED -> IP, IP)
  ed_discharged_home <- sum(df$Patient_Type_Title == "ED Visit")
  percent_ed_discharged_home <- (ed_discharged_home / nrow(df)) * 100
  ed_admitted <- sum(df$Patient_Type_Title != "ED Visit" & df$ED_entry == 1)
  percent_ed_admitted <- (ed_admitted / nrow(df)) * 100
  admitted_without_ed <- sum(df$Patient_Type_Title != "ED Visit" & df$ED_entry == 0)
  percent_admitted_without_ed <- (admitted_without_ed / nrow(df)) * 100
  
  # patient type df
  ptype_counts <- list(
    ED_Discharged_Home_Count = ed_discharged_home,
    ED_Admitted_Count = ed_admitted,
    Admitted_Without_ED_Count = admitted_without_ed)
  
  # print results
  cat("ED Discharged Home:\n")
  cat("Count:", ed_discharged_home, "\n")
  cat("Percentage:", percent_ed_discharged_home, "%\n\n")
  cat("ED Admitted:\n")
  cat("Count:", ed_admitted, "\n")
  cat("Percentage:", percent_ed_admitted, "%\n\n")
  cat("Admitted Without ED:\n")
  cat("Count:", admitted_without_ed, "\n")
  cat("Percentage:", percent_admitted_without_ed, "%\n\n")
  
  # median (IQR) age, in months
  df_summary <- df %>%
    summarize(
      Median_Age_Months = round(median(Admit_Age_In_Days) / 30.44, 2),
      Q1 = round(quantile(Admit_Age_In_Days, 0.25) / 30.44, 2),
      Q3 = round(quantile(Admit_Age_In_Days, 0.75) / 30.44, 2))
  
  # print age statistics
  cat("Admit_Age_In_Mo - Median:", df_summary$Median_Age_Months, "months\n")
  cat("Admit_Age_In_Mo - IQR (Q1 - Q3):", df_summary$Q1, "months -", df_summary$Q3, "months\n")
  
  # report percent in each age category
  age_groups = table(df$Age_Group)
  age_percents <- round(table(df$Age_Group) / nrow(df) * 100, 2)
  cat("Admit_Age_In_Days - Percent:\n")
  print(age_percents)
  
  # report percent male
  df_gender_summary <- df %>%
    summarize(Percent_Male = round(sum(Gender_Title == "Male") / n() * 100, 2))
  cat("Gender_Title - Percent Male:", df_gender_summary$Percent_Male, "%\n")
  
  # report percent in each ethnicity category
  Ethnicity_Summary <- df %>%
    summarize(
      Percent_Hispanic = round(sum(Ethnicity_Title == "Hispanic or Latino") / n() * 100, 2),
      Percent_Not_Hispanic = round(sum(Ethnicity_Title == "Not Hispanic or Latino") / n() * 100, 2),
      Percent_Unknown = round(sum(Ethnicity_Title == "Unknown") / n() * 100, 2))
  cat("Ethnicity_Title - Percent Hispanic or Latino:", Ethnicity_Summary$Percent_Hispanic, "%\n")
  cat("Ethnicity_Title - Percent Not Hispanic or Latino:", Ethnicity_Summary$Percent_Not_Hispanic, "%\n")
  cat("Ethnicity_Title - Percent Unknown:", Ethnicity_Summary$Percent_Unknown, "%\n")
  
  # report percent in each racial group
  race_columns <- c("Race_White", "Race_Black", "Race_Asian", "Race_Pacific_Islander", "Race_American_Indian", "Race_Other")
  # column for multiple races
  df <- df %>%
    mutate(Multiple = ifelse(rowSums(select(., all_of(race_columns)) == "Y") > 1, "Y", "N"))
  # percent in each individual category
  Race_Summary <- df %>%
    summarize(across(all_of(race_columns), ~ round(sum(. == "Y" & Multiple == "N") / n() * 100, 2))) %>%
    rename_with(~ paste0(.x, "_Percentage"), all_of(race_columns))
  # percent in multiple categories
  Multiple_Percentage <- df %>%
    summarize(Multiple_Percentage = round(sum(Multiple == "Y") / n() * 100, 2))
  # percent unknown
  Unknown_Percentage <- 100 - sum(Race_Summary) - Multiple_Percentage
  # print race category percents
  Race_Summary <- Race_Summary %>%
    bind_rows(data.frame(Multiple = Multiple_Percentage, Unknown = Unknown_Percentage)) %>%
    set_names(nm = c("Race_White_Percentage", "Race_Black_Percentage", "Race_Asian_Percentage", 
                     "Race_Pacific_Islander_Percentage", "Race_American_Indian_Percentage", 
                     "Race_Other_Percentage", "Multiple_Percentage", "Unknown_Percentage"))
  print(Race_Summary)
  # save counts for stat testing
  Race_Summary <- df %>%
    summarize(across(all_of(c(race_columns, "Multiple")), ~ sum(. == "Y")))
  
  # report test categories
  test_columns <- c("culture", "PCR", "SC2_multi", "antigen", "DNA_probe", "unspecified")
  # column for multiple tests
  df <- df %>%
    mutate(Multiple = ifelse(rowSums(select(., all_of(test_columns)) > 0) > 1, 1, 0))
  # percent in each individual category
  Test_Summary <- df %>%
    summarize(across(all_of(test_columns), ~ round(sum(. > 0 & Multiple == 0) / n() * 100, 3))) %>%
    rename_with(~ paste0(.x, "_Percentage"), all_of(test_columns))
  # percent in multiple categories
  Multiple_Percentage <- df %>%
    summarize(Multiple_Percentage = round(sum(Multiple == 1) / n() * 100, 3))
  # percent without test
  No_Test_Percentage <- 100 - sum(Test_Summary) - Multiple_Percentage
  # print test category percents
  Test_Summary <- Test_Summary %>%
    bind_rows(data.frame(Multiple = Multiple_Percentage, No_Test = No_Test_Percentage)) %>%
    set_names(nm = c("culture", "PCR", "SC2_multi", "antigen", "DNA_probe", 
                     "unspecified", "Multiple", "No_Test"))
  print(Test_Summary)
  # save counts for stat testing
  test_columns <- c("culture", "PCR", "SC2_multi", "antigen", "DNA_probe")
  Test_Summary <- df %>%
    summarize(across(all_of(c(test_columns, "Multiple")), ~ sum(. > 0)),
              No_Test = sum(rowSums(select(., all_of(test_columns))) == 0))
  
  # print percent with CCC
  summarize_and_print_column(df, "Complex_Chronic_Condition_Flag")
  
  # print percent with neonatal CCC
  summarize_and_print_column(df, "Premature_And_Neonatal_Flag")
  
  # print percent with CCC per age group
  complex_condition_counts <- table(df$Age_Group[df$Complex_Chronic_Condition_Flag == "Y"])
  complex_condition_percentages <- round(complex_condition_counts / table(df$Age_Group) * 100, 2)
  cat("Age Categories - Complex Chronic Condition Percentage:\n")
  print(complex_condition_percentages)
  
  # print percent with positive test per age group
  if ("pos" %in% colnames(df)) {
    summarize_and_print_column(df, "pos")
    pos_counts_by_age <- df %>%
      filter(!is.na(Age_Group)) %>%
      group_by(Age_Group) %>%
      summarize(
        Pos_1_Count = sum(pos == 1, na.rm = TRUE),
        Pos_0_Count = sum(pos == 0, na.rm = TRUE),
        Pos_1_Percent = round(mean(pos == 1, na.rm = TRUE) * 100, 5))
    cat("Age Groups - Percentage of rows where pos == 1:\n")
    print(data.frame(pos_counts_by_age))
    
    # report percent of positive tests by test type
    test_columns <- c("culture", "antigen", "PCR", "SC2_multi", "DNA_probe")
    if(all(test_columns %in% colnames(df))) {
      test_counts <- lapply(test_columns, function(test_col) {
        df %>%
          filter(.data[[test_col]] > 0) %>%
          summarise(
            Pos_1_Count = sum(.data[[test_col]] > 0 & pos == 1, na.rm = TRUE),
            Pos_0_Count = sum(.data[[test_col]] > 0 & pos == 0, na.rm = TRUE),
            Pos_1_Percent = round(sum(.data[[test_col]] > 0 & pos == 1) / sum(.data[[test_col]] > 0) * 100, 4)) %>%
          mutate(Test = test_col)})
      test_counts <- bind_rows(test_counts)
      cat("Test Counts - Positive cases where test > 0 and pos == 1 or pos == 0:\n")
      print(test_counts)}}
  else {
    pos_counts_by_age = c()
    test_counts = c()}
  
  # statistics of interest only for patients diagnosed with RSV
  if ("flu" %in% colnames(df)) {
    # print percent with concurrent influenza
    summarize_and_print_column(df, "flu")
    
    # print percent with concurrent SC2
    summarize_and_print_column(df, "sc2")
    
    # print percent admitted from ED
    ED_Dispo_Admitted_Percent <- df %>%
    filter(ED_entry == 1) %>%
    filter(!is.na(ED_Dispo)) %>%
    summarize(ED_Dispo_Admitted_Percent = round(sum(ED_Dispo == "ED_Admission") / n() * 100, 2))
    cat("Percent ED Patients Admitted:", ED_Dispo_Admitted_Percent$ED_Dispo_Admitted_Percent, "%\n")
    
    # only consider inpatients for the following statistics
    # print median (IQR) length of stay
    df = df %>% filter %>%
      filter(Patient_Type_Title != "ED Visit")
    LOS_Summary <- df %>%
      summarize(Median_Length_Of_Stay = round(median(Length_Of_Stay), 2),
      Q1_Length_Of_Stay = round(quantile(Length_Of_Stay, 0.25), 2),
      Q3_Length_Of_Stay = round(quantile(Length_Of_Stay, 0.75), 2))
    cat("Length_Of_Stay - Median:", LOS_Summary$Median_Length_Of_Stay, "\n")
    cat("Length_Of_Stay - Q1:", LOS_Summary$Q1_Length_Of_Stay, "\n")
    cat("Length_Of_Stay - Q3:", LOS_Summary$Q3_Length_Of_Stay, "\n")
    
    # print percent in each LOS category
    los_groups <- c(0, 1, 2, 4, 7, Inf)
    los_labels <- c("0-1 days", "1-2 days", "2-4 days", "5-7 days", ">7 days")
    df$LOS_Group <- cut(df$Length_Of_Stay, breaks = los_groups, labels = los_labels, right = FALSE)
    los_groups = table(df$LOS_Group)
    los_percents <- round(table(df$LOS_Group) / nrow(df) * 100, 2)
    cat("Length_Of_Stay - Percent:\n")
    print(los_percents)

    # print percent of IPs admitted to intensive care
    df$nicu_icu = ifelse(df$ICU_Flag == "Y" | df$NICU_Flag == "Y", "Y", "N")
    summarize_and_print_column(df, "nicu_icu")
    
    # print percent of IPs requiring mechanical ventilation
    summarize_and_print_column(df, "Mechanical_Vent_Flag")
  
    # print percent of IPs requiring ECMO
    summarize_and_print_column(df, "ECMO_Flag")
    
    # print percent of IPs expiring in hospital
    summarize_and_print_column(df, "Discharge_Mortality_Flag")}
    
    else {
      los_groups = c()}
  
  return(list(ptype = ptype_counts, age = age_groups, los = los_groups, race = Race_Summary, test = Test_Summary, pos_age = pos_counts_by_age, pos_test = test_counts))}
```

This function takes two dfs with column "Age_Group" (character) as input. It determines the distribution of the categorical variable Age_Group in the reference_df and randomly samples rows in target_df, with replacement, to generate a new target_df with the same distribution of Age_Group as reference_df. 

```{r}
bootstrap_df_by_age <- function(target_df, reference_df, sample_size = nrow(target_df)) {
  # calc the distribution of Age_Group in reference_df
  reference_distribution <- table(reference_df$Age_Group)
  subset_df <- data.frame()
  # loop through each age group
  for (age_group in unique(reference_df$Age_Group)) {
    # calc the target percent for this age group
    target_percentage <- reference_distribution[age_group] / sum(reference_distribution)
    # calc the sample size for this age group
    n_to_select <- round(target_percentage * sample_size)
    # select rows from target_df matching the age group
    available_rows <- target_df %>%
      filter(Age_Group == age_group)
    # make sure there are rows in this age group present
    if (nrow(available_rows) > 0) {
      # randomly sample rows with replacement
      sampled_rows <- available_rows %>% sample_n(size = n_to_select, replace = TRUE)
      subset_df <- bind_rows(subset_df, sampled_rows)}}
  return(subset_df)}
```

Add the columns "Age_Group", "nicu_icu", and "Race" to the dfs with diagnosed and tested patient information.

```{r}
# create the "Age_Group" column from the Admit_Age_In_Days column
pt = add_age_group(pt)
ptest = add_age_group(ptest)

# create the "nicu_icu" column from the binary ICU flags
pt$nicu_icu = ifelse(pt$ICU_Flag == "Y" | pt$NICU_Flag == "Y", "Y", "N")

# create the "Race" column from the binary race columns
race_columns <- c("Race_White", "Race_Black", "Race_Asian", "Race_Pacific_Islander", "Race_American_Indian", "Race_Other")
pt$Race <- apply(pt[race_columns], 1, function(row) {
  y_columns <- names(row[row == "Y"])
  if (length(y_columns) == 1) {
    return(y_columns)
  } else if (length(y_columns) > 1) {
    return("Multiple")
  } else {
    return("Unknown")}})
ptest$Race <- apply(ptest[race_columns], 1, function(row) {
  y_columns <- names(row[row == "Y"])
  if (length(y_columns) == 1) {
    return(y_columns)
  } else if (length(y_columns) > 1) {
    return("Multiple")
  } else {
    return("Unknown")}})
```

Report the median (IQR) of monthly RSV cases across all hospitals during the intermediate period.

```{r}
# print median (IQR) of cases between phases
pandemic <- pt %>%
  filter(Date >= as.Date('2020-04-01') & Date < as.Date('2021-04-01')) %>%
  mutate(YearMonth = format(Date, "%Y-%m")) %>%
  group_by(YearMonth) %>%
  summarise(Count = n())

print(paste("Median:", median(pandemic$Count)))
print(paste("Q1:", quantile(pandemic$Count, probs = 0.25)))
print(paste("Q3:", quantile(pandemic$Count, probs = 0.75)))
rm(pandemic)
```

Produce summary statistics for patients tested for RSV in the pre-pandemic and post-emergence phases, and identify statistically significant differences in demographic or clinical variables across the two phases (Table 1).

```{r}
# pre-pandemic
pre_ptest = subset(ptest, Date < as.Date("2020-04-01"))
print("Pre-pandemic tested patients:")
pre_ptest_cat = analyze_dataframe(pre_ptest)

# post-emergence
post_ptest <- subset(ptest, Date >= as.Date("2021-04-01"))
print("Post-emergence tested patients:")
post_ptest_cat = analyze_dataframe(post_ptest)

# compare pre-pandemic and post-emergence phases
agepv = wilcox.test(pre_ptest$Admit_Age_In_Days/30.44, post_ptest$Admit_Age_In_Days/30.44, paired = FALSE)$p.value
print("Age p-value:")
agepv
agecatpv = chisq.test(rbind(pre_ptest_cat$age, post_ptest_cat$age))$p.value
print("Age group p-value:")
agecatpv
sexpv = fisher.test(rbind(table(pre_ptest$Gender_Title)[1:2], table(post_ptest$Gender_Title)[1:2]))$p.value
print("Sex p-value:")
sexpv
ethnpv = fisher.test(rbind(table(pre_ptest$Ethnicity_Title)[1:2], table(post_ptest$Ethnicity_Title)[1:2]))$p.value
print("Ethnicity p-value:")
ethnpv
racepv = chisq.test(rbind(pre_ptest_cat$race, post_ptest_cat$race))$p.value
print("Race p-value:")
racepv
cccpv = fisher.test(rbind(table(pre_ptest$Complex_Chronic_Condition_Flag[pre_ptest$Complex_Chronic_Condition_Flag != "P"]), 
                          table(post_ptest$Complex_Chronic_Condition_Flag[post_ptest$Complex_Chronic_Condition_Flag != "P"])))$p.value
print("CCC p-value:")
cccpv

# Fisher's for each age group's CCC distribution
cc_pv <- list()
age_groups <- unique(as.character(ptest$Age_Group))
for (age_group in age_groups) {
  table_pre <- table(pre_ptest$Complex_Chronic_Condition_Flag[pre_ptest$Age_Group == age_group & pre_ptest$Complex_Chronic_Condition_Flag != "P"])
  table_post <- table(post_ptest$Complex_Chronic_Condition_Flag[post_ptest$Age_Group == age_group & post_ptest$Complex_Chronic_Condition_Flag != "P"])
  pvalue <- fisher.test(rbind(table_pre, table_post))$p.value
  cc_pv[[age_group]] <- pvalue}
# print results
for (i in 1:length(age_groups)) {
  cat("Age Group:", age_groups[i], "- p-value:", cc_pv[[i]], "\n")}

prempv = fisher.test(rbind(table(pre_ptest$Premature_And_Neonatal_Flag[pre_ptest$Premature_And_Neonatal_Flag %in% c("N", "Y")]),                                  table(post_ptest$Premature_And_Neonatal_Flag[pre_ptest$Premature_And_Neonatal_Flag %in% c("N", "Y")])))$p.value
print("Premature & neonatal CCC p-value:")
prempv

ptypepv = chisq.test(rbind(unlist(pre_ptest_cat$ptype), unlist(post_ptest_cat$ptype)))$p.value
print("Patient type p-value:")
ptypepv

testpv = chisq.test(rbind(unlist(pre_ptest_cat$test), unlist(post_ptest_cat$test)))$p.value
print("Test type p-value:")
testpv

# Fisher's for each age group's test positivity
pos_pv <- list()
age_groups <- unique(as.character(ptest$Age_Group))
for (age_group in age_groups) {
  table_pre <- table(pre_ptest$pos[pre_ptest$Age_Group == age_group])
  table_post <- table(post_ptest$pos[post_ptest$Age_Group == age_group])
  pvalue <- fisher.test(rbind(table_pre, table_post))$p.value
  pos_pv[[age_group]] <- pvalue}
for (i in 1:length(age_groups)) {
  cat("Age stratified test positivity:", age_groups[i], " p-value:", pos_pv[[i]], "\n")}

pos_pv = fisher.test(rbind(table(pre_ptest$pos), table(post_ptest$pos)))$p.value
print("Positive test p-value:")
pos_pv
pos_cult_pv = fisher.test(rbind(table(pre_ptest$pos[pre_ptest$culture > 0]), table(post_ptest$pos[post_ptest$culture > 0])))$p.value
print("Positive culture p-value:")
pos_cult_pv
pos_ant_pv = fisher.test(rbind(table(pre_ptest$pos[pre_ptest$antigen > 0]), table(post_ptest$pos[post_ptest$antigen > 0])))$p.value
print("Positive antigen p-value:")
pos_ant_pv
pos_pcr_pv = fisher.test(rbind(table(pre_ptest$pos[pre_ptest$PCR > 0]), table(post_ptest$pos[post_ptest$PCR > 0])))$p.value
print("Positive PCR p-value:")
pos_pcr_pv
pos_dna_pv = fisher.test(rbind(table(pre_ptest$pos[pre_ptest$DNA_probe > 0]), table(post_ptest$pos[post_ptest$DNA_probe > 0])))$p.value
print("Positive DNA probe p-value:")
pos_dna_pv

rm(list = ls(pattern = "pv|tab"))
rm(i)
```

Produce summary statistics for patients diagnosed with RSV in the pre-pandemic and post-emergence phases, and identify statistically significant differences in demographic or clinical variables across the two phases (Table 2).

```{r}
pt = pt %>% filter(Date < as.Date("2020-04-01") | Date >= as.Date("2021-04-01"))
pt$phase = ifelse(pt$Date < as.Date("2020-04-01"), "Pre-Pandemic", "Post-Emergence")

# pre-pandemic
pre_pt = subset(pt, Date < as.Date("2020-04-01"))
print("Pre-pandemic diagnosed patients:")
pre_pt_cat = analyze_dataframe(pre_pt)

# post-emergence
post_pt <- subset(pt, Date >= as.Date("2021-04-01"))
print("Post-emergence diagnosed patients:")
post_pt_cat = analyze_dataframe(post_pt)

# compare pre-pandemic and post-emergence phases
agepv = wilcox.test(pre_pt$Admit_Age_In_Days/30.44, post_pt$Admit_Age_In_Days/30.44, paired = FALSE)$p.value
print("Age p-value:")
agepv
agecatpv = chisq.test(rbind(pre_pt_cat$age, post_pt_cat$age))$p.value
print("Age group p-value:")
agecatpv
sexpv = fisher.test(rbind(table(pre_pt$Gender_Title)[1:2], table(post_pt$Gender_Title)[1:2]))$p.value
print("Sex p-value:")
sexpv
ethnpv = fisher.test(rbind(table(pre_pt$Ethnicity_Title)[1:2], table(post_pt$Ethnicity_Title)[1:2]))$p.value
print("Ethnicity p-value:")
ethnpv
racepv = chisq.test(rbind(pre_pt_cat$race, post_pt_cat$race))$p.value
print("Race p-value:")
racepv
flupv = fisher.test(rbind(table(pre_pt$flu), table(post_pt$flu)))$p.value
print("Flu p-value:")
flupv
print("SC2 p-value:")
sc2pv = fisher.test(rbind(table(pre_pt$sc2), table(post_pt$sc2)))$p.value
sc2pv
cccpv = fisher.test(rbind(table(pre_pt$Complex_Chronic_Condition_Flag), table(post_pt$Complex_Chronic_Condition_Flag)))$p.value
print("CCC p-value:")
cccpv

# Fisher's for each age group's CCC distribution
cc_pv <- list()
age_groups <- unique(as.character(pt$Age_Group))
for (age_group in age_groups) {
  table_pre <- table(pre_pt$Complex_Chronic_Condition_Flag[pre_pt$Age_Group == age_group])
  table_post <- table(post_pt$Complex_Chronic_Condition_Flag[post_pt$Age_Group == age_group])
  pvalue <- fisher.test(rbind(table_pre, table_post))$p.value
  cc_pv[[age_group]] <- pvalue}
for (i in 1:length(age_groups)) {
  cat("Age stratified CCC:", age_groups[i], "p-value:", cc_pv[[i]], "\n")}

prempv = fisher.test(rbind(table(pre_pt$Premature_And_Neonatal_Flag), table(post_pt$Premature_And_Neonatal_Flag)))$p.value
print("Premature & neonatal CCC p-value:")
prempv
ptypepv = chisq.test(rbind(unlist(pre_pt_cat$ptype), unlist(post_pt_cat$ptype)))$p.value
print("Patient type p-value:")
ptypepv
testpv = chisq.test(rbind(unlist(pre_pt_cat$test), unlist(post_pt_cat$test)))$p.value
print("Test type p-value:")
testpv
admpv = fisher.test(rbind(table(pre_pt$ED_Dispo[pre_pt$ED_entry == 1]), 
                          table(post_pt$ED_Dispo[post_pt$ED_entry == 1])))$p.value
print("ED admission p-value:")
admpv
lospv = wilcox.test(pre_pt$Length_Of_Stay[pre_pt$Patient_Type_Title != "ED Visit"], 
                    post_pt$Length_Of_Stay[post_pt$Patient_Type_Title != "ED Visit"], paired = FALSE)$p.value
print("Length of stay p-value:")
lospv
loscatpv = chisq.test(rbind(pre_pt_cat$los, post_pt_cat$los))$p.value
print("Length of stay (categorical) p-value:")
loscatpv
icupv = fisher.test(rbind(table(pre_pt$nicu_icu[pre_pt$Patient_Type_Title != "ED Visit"]), 
                          table(post_pt$nicu_icu[post_pt$Patient_Type_Title != "ED Visit"])))$p.value
print("ICU p-value:")
icupv
mechpv = fisher.test(rbind(table(pre_pt$Mechanical_Vent_Flag[pre_pt$Patient_Type_Title != "ED Visit"]), 
                           table(post_pt$Mechanical_Vent_Flag[post_pt$Patient_Type_Title != "ED Visit"])))$p.value
print("Mechanical ventilation p-value:")
mechpv
ecmopv = fisher.test(rbind(table(pre_pt$ECMO_Flag[pre_pt$Patient_Type_Title != "ED Visit"]), 
                           table(post_pt$ECMO_Flag[post_pt$Patient_Type_Title != "ED Visit"])))$p.value
print("ECMO p-value:")
ecmopv
deathpv = fisher.test(rbind(table(pre_pt$Discharge_Mortality_Flag[pre_pt$Patient_Type_Title != "ED Visit"]), 
                            table(post_pt$Discharge_Mortality_Flag[post_pt$Patient_Type_Title != "ED Visit"])))$p.value
print("Death p-value:")
deathpv

rm(list = ls(pattern = "pv|tab"))
rm(i)
```

Plot the proportion of tested patients in each age group (Figure 2A). 

```{r}
test_age = rbind(pre_ptest_cat$age/nrow(pre_ptest), post_ptest_cat$age/nrow(post_ptest))
rownames(test_age) <- c("Pre-Pandemic", "Post-Emergence")

# pivot the data to long format for plotting
test_age_long = gather(as.data.frame(test_age), key = "Age Group", value = "Proportion")
test_age_long$Group <- rownames(test_age)
test_age_long$Group <- factor(test_age_long$Group, levels = c("Pre-Pandemic", "Post-Emergence"))
test_age_long$`Age Group` <- factor(test_age_long$`Age Group`, levels = c("0-3 months", 
                             "3-6 months", "6-12 months", "1-2 years", "2-4 years", "5-17 years"))

# plot the proportions
fig2a <- ggplot(test_age_long, aes(x = `Age Group`, y = Proportion, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "",
       x = "Age Group",
       y = "Proportion of Tests") +
  scale_fill_manual(values = c("grey", "darkmagenta")) + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        legend.position = "top",
        text = element_text(size = 12)) +
  guides(fill = guide_legend(title = NULL))
fig2a
rm(test_age, test_age_long)
```

Use bootstrapping to generate a post-emergence testing df with the age distribution (in 90-day bins) derived from the pre-pandemic testing df. Plot the pre-pandemic, post-emergence, and test-adjusted age distributions of tested patients (Supplementary Figure 3).

```{r}
# match testing by 3-month periods of age
post_ptest$Age_Group <- as.character(floor(post_ptest$Admit_Age_In_Days / 90))
pre_ptest$Age_Group <- as.character(floor(pre_ptest$Admit_Age_In_Days / 90))

# bootstrapping post-emergence phase tests
boot_post = bootstrap_df_by_age(post_ptest, pre_ptest)

# plot the pre-pandemic, post-emergence, and test-adjusted age distributions
age_tests <- rbind(data.frame(Age = pre_ptest$Admit_Age_In_Days/365, Group = "Pre-Pandemic"),
                   data.frame(Age = post_ptest$Admit_Age_In_Days/365, Group = "Post-Emergence"),
                   data.frame(Age = boot_post$Admit_Age_In_Days/365, Group = "Post-Emergence\n(Test Adjusted)"))
age_tests$Group = factor(age_tests$Group, levels = c("Pre-Pandemic", "Post-Emergence", "Post-Emergence\n(Test Adjusted)"))

sf3 <- ggplot(age_tests, aes(x = Group, y = Age)) +
  geom_boxplot(outlier.shape = NA, fill = "grey") +
  labs(x = "", y = "Age of Tested Individuals (Years)") +
  theme_classic() +
  coord_cartesian(ylim = c(0, 19)) +
  geom_signif(comparisons = list(c("Pre-Pandemic","Post-Emergence"), 
                                 c("Pre-Pandemic", "Post-Emergence\n(Test Adjusted)"),
                                 c("Post-Emergence", "Post-Emergence\n(Test Adjusted)")),
              test = "wilcox.test",
              map_signif_level = TRUE,
              textsize = 3, 
              y_position = (c(17, 18.5, 17.5)),
              tip_length = 0.005) +
  scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18)) +
  theme(legend.position = "none",
        text = element_text(size = 12))
sf3
ggsave("figs/suppfig3.pdf", plot = sf3, width = 6, height = 6)
rm(age_tests, sf3)
```

Assess whether the distribution of patients diagnosed with RSV would have changed if the age of patients tested for RSV remained constant from the pre-pandemic to the post-emergence phase. Conduct 50 replicates of bootstrapping to ensure that the results are robust to sampling error.

```{r}
# create empty df to store results
age_sim <- data.frame(min = numeric(0), q1 = numeric(0), median = numeric(0), 
                      q3 = numeric(0), max = numeric(0))
for (i in 1:50){
  # bootstrap tests according to pre-pandemic age distribution
  boot_post = bootstrap_df_by_age(post_ptest, pre_ptest)
  # isolate positive bootstrapped tests
  pt_post = boot_post %>% filter(Discharge_ID %in% pt$Discharge_ID)
  # append age range to df
  age_sim[nrow(age_sim)+1, ] <- c(quantile(pt_post$Admit_Age_In_Days))}

# check for consistency across replicates
summary(age_sim/30.44) # convert to months
rm(age_sim)
```

Plot the pre-pandemic, post-emergence, and test-adjusted age distributions of patients with a laboratory-confirmed diagnosis of RSV (Figure 2B).

```{r}
# create plotting df
age_adj_df <- rbind(data.frame(Age = (pre_pt %>% filter(Num_Tests > 0))$Admit_Age_In_Days/365, Group = "Pre-Pandemic"),
                    data.frame(Age = (post_pt %>% filter(Num_Tests > 0))$Admit_Age_In_Day/365, Group = "Post-Emergence"),
                    data.frame(Age = pt_post$Admit_Age_In_Day/365, Group = "Post-Emergence\n(Test Adjusted)"))
age_adj_df$Group = factor(age_adj_df$Group, levels = c("Pre-Pandemic", "Post-Emergence", "Post-Emergence\n(Test Adjusted)"))

# print summary statistics for each group
tapply(12*age_adj_df$Age, age_adj_df$Group, quantile) # convert to months

# plot age distributions
fig2b <- ggplot(age_adj_df, aes(x = Group, y = Age)) +
  geom_boxplot(outlier.shape = NA, fill = "grey") +
  labs(x = "", y = "Age of Diagnosed Individuals (Years)") +
  theme_classic() +
  coord_cartesian(ylim = c(0, 7)) +
  geom_signif(comparisons = list(c("Pre-Pandemic","Post-Emergence"), 
                                 c("Pre-Pandemic", "Post-Emergence\n(Test Adjusted)"),
                                 c("Post-Emergence", "Post-Emergence\n(Test Adjusted)")),
              test = "wilcox.test",
              map_signif_level = TRUE,
              textsize = 3, 
              y_position = c(5.1,5.5,6.0),
              tip_length = 0.003) +
  theme(legend.position = "none",
        text = element_text(size = 12))
fig2b
rm(age_adj_df)
```

Generate Figure 2.

```{r}
fig2 = plot_grid(fig2a, fig2b, nrow = 1, labels = c("A", "B"))
fig2
ggsave("figs/fig2.pdf", plot = fig2, width = 11, height = 5)
```

